Tesseroids: forward modeling gravitational fields in spherical coordinates

### 2014-11-15 manuscript.tex
Algorithm description done
Need to tie it with the calc_tess_model_adapt function

\documentclass[paper,twocolumn]{geophysics}
%\documentclass[manuscript]{geophysics}
%\documentclass[manuscript,endfloat]{geophysics}
\usepackage{graphicx}
\newcommand{\mitbf}[1]{\mathbf{{#1}}}

@@

\title{Title}
\lefthead{Meh}
\righthead{Meh}
% manuscript number
\ms{Submission}

\author{
    Leonardo Uieda\footnotemark[1]\footnotemark[2]
    Val\'eria C. F. Barbosa\footnotemark[2],
    Vanderlei C. Oliveira Jr\footnotemark[2],
    and
    Carla Braitenberg\footnotemark[3]
}

@@

These functions require three \texttt{GLQ} structures with roots scaled to the
integration limits $[\lambda_1, \lambda_2], [\phi_1, \phi_2], [r_1, r_2]$.

@@

Better performance is achieved by moving some trigonometric function
evaluations to the outer loops
and precomputing the sine and cossine of latitudes.

@@

\begin{figure}
    \centering
    \includegraphics{figs/tesseroid-split}
    \caption{
        Adaptive discretization
        of the tesseroid shown in (a)
        for a computation point P
        using the distance-size ratio $D$ equal to
        (b) 1, (c) 2, and (d) 6.
        $d$ is the Cartesian distance between
        the center of the top face of the tesseroid
        and $P$.
        $L_r$, $L_\phi$, and $L_\lambda$ are the dimensions of the tesseroid.
        Note that increasing $D$
        results in a fine division of the tesseroid
        close the computation point
        and a coarser division further away.
    }
    \label{fig:ratio}
\end{figure}

>>>>

\documentclass[paper,twocolumn]{geophysics}
%\documentclass[manuscript]{geophysics}
%\documentclass[manuscript,endfloat]{geophysics}
\usepackage{graphicx}
\usepackage{amsmath}
\newcommand{\mitbf}[1]{\mathbf{{#1}}}

@@

\title{
Tesseroids: forward modeling gravitational fields in spherical coordinates
}
\righthead{Tesseroids}
% manuscript number
\ms{Submission}

\author{
    Leonardo Uieda\footnotemark[1]\footnotemark[2],
    Val\'eria C. F. Barbosa\footnotemark[2],
    Carla Braitenberg\footnotemark[3],
    and
    Vanderlei C. Oliveira Jr\footnotemark[2]
}

@@

These functions require three \texttt{GLQ} structures with roots scaled to the
integration limits $[\lambda_1, \lambda_2], [\phi_1, \phi_2], [r_1, r_2]$
(see equations~\ref{eq:tesspot}-\ref{eq:tesstensor}).

@@

Better performance is achieved by
precomputing the sine and cossine of latitudes
and moving some trigonometric function
evaluations to the outer loops.

@@

The adaptive discretization algorithm of \citet{Li2011} recursively
divides a tesseroid.

A tesseroid will be divided when the smallest distance between it and the
computation point is smaller than a constant times the largest dimension of the
tesseroid.

In our implementation,
we measure the distance between
the computation point
and the center of the top face tesseroid
(see Figure~\ref{fig:division}a).

The distance $d$ is

\begin{equation}
    d = \left[
        r^2 + r_2^2 - 2 r r_2
        \cos(\sin\phi\sin\Delta\phi + \cos\phi\cos\Delta\phi
            \cos(\lambda - \Delta\lambda))
        \right],
    \label{eq:distance}
\end{equation}

where
$(r, \lambda, \phi)$ are the coordinates of the computation point,
$\Delta\lambda = 0.5(\lambda_2 + \lambda_1)$,
and
$\Delta\phi = 0.5(\phi_2 + \phi_1)$.

We define the dimensions of the tesseroid
along longitude, latitude, and radius, respectively, as

\begin{equation}
    L_\lambda = r_2 \arccos(\sin^2\Delta\phi +
        \cos^2\Delta\phi\cos(\lambda_2 - \lambda_1)),
    \label{eq:sizelon}
\end{equation}
\begin{equation}
    L_\phi = r_2 \arccos(\sin\phi_2\sin\phi_1 + \cos\phi_2\cos\phi_1),
\end{equation}
\begin{equation}
    L_r = r_2 - r_2.
    \label{eq:sizer}
\end{equation}

We fix the GLQ order $N=2$ for all three dimensions, resulting in eight point
masses in total.

The criterion we use to determine if a tesseroid must be divided is

\begin{equation}
    \frac{d}{L_i} < D,
    \label{eq:condition}
\end{equation}
\noindent
where $i \in (\lambda, \phi, r)$
and $D$ is a positive scalar called the ``distance-size ratio''.
$D$ determines how close the
computation point can be before we must divide the tesseroid.

We used a stack-based implementation of the algorithm instead of using
recursive function calls.

Stacks are array-like data structures with a particular way of inserting and
removing elements from it.

In a stack, one can only insert elements on the top (the last empty position).

Likewise,
one can only remove the last element that was placed on the stack
(commonly referred to as ``popping'' the stack).

Because of this stacks are also known as ``Last-in-first-out'', or LIFO,
structures.

LIFOs are often described as being analogous to a stack of plates.

One can only place a new plate on the top of the stack and
can only remove a plate safely from the top of the stack.

The algorithm is described for a single tesseroid and computation point.

The recursive division algorithm implemented with a stack is as follows.

Start by initializing the stack with the tesseroid.

Thus, so far, the stack has only one element.

Once the stack is initialized, the steps of the algorithm are:

\begin{enumerate}
    \item ``Pop'' the stack (i.e., take the last element from it).
    \item Compute the distance $d$ between the tesseroid and
        the computation point (equation~\ref{eq:distance}).
    \item Compute the dimensions of the tesseroid $L_\lambda$, $L_\phi$,
        and $L_r$ (equations~\ref{eq:sizelon}-\ref{eq:sizer}).
    \item Check the condition in equation~\ref{eq:condition} for each
        dimension of the tesseroid.
    \item If none of the dimensions fail the above condition,
        compute the gravitational effect of the tesseroid using the
        Gauss-Legendre Quadrature with order two
        (equation~\ref{eq:glq3d}).
    \item If any of the dimensions fail the condition,
        divide the tesseroid in half along each dimension that failed the
        condition.
        Then, place the smaller tesseroids into the stack.
    \item Repeat the above steps until the stack is empty.
\end{enumerate}

\begin{figure}
    \centering
    \includegraphics{figs/tesseroid-split}
    \caption{
        Adaptive discretization
        of the tesseroid shown in (a)
        for a computation point P
        using the distance-size ratio $D$ equal to
        (b) 1, (c) 2, and (d) 6.
        $d$ is the Cartesian distance between
        the center of the top face of the tesseroid
        and $P$.
        $L_r$, $L_\phi$, and $L_\lambda$ are the dimensions of the tesseroid.
        Note that increasing $D$
        results in a fine division of the tesseroid
        close the computation point
        and a coarser division further away.
    }
    \label{fig:division}
\end{figure}
